\proglang{collapse}: Advanced and Fast Statistical Computing and Data Transformation in \proglang{R}

### 2025-05-05 misc/JSS article/final/article.tex
Run benchmark on M4 Mac.

%% For Sweave-based articles about R packages:
%% need no \usepackage{Sweave}
%% need no \usepackage{Sweave.sty}
\SweaveOpts{engine=R, eps=FALSE, keep.source = TRUE, cache=FALSE}
% \renewenvironment{knitrout}{\setlength{\topsep}{0mm}}{}

<<preliminaries, echo=FALSE, message=FALSE, warning=FALSE, results=hide>>=
# knitr::opts_chunk$set(cache = FALSE, size = "small")
rm(list = ls()); gc()
options(prompt = "R> ", continue = "+  ", width = 77, digits = 4, useFancyQuotes = FALSE, warn = 1)

# Loading libraries and installing if unavailable
if(!requireNamespace("fastverse", quietly = TRUE)) install.packages("fastverse")
library(fastverse) # loads data.table, collapse, magrittr and kit (not used)
# Package versions used in the article:
# fastverse 0.3.4, collapse 2.1.2, data.table 1.16.4,
# bench 1.1.3, dplyr 1.1.4, tidyr 1.3.1, matrixStats 1.0.0

# Reset collapse options (if set)
set_collapse(nthreads = 1L, remove = NULL, stable.algo = TRUE, sort = TRUE,
             digits = 2L, stub = TRUE, verbose = 1L, mask = NULL, na.rm = TRUE)

# For presentation: removing whitespace around labels
vlabels(GGDC10S) <- trimws(vlabels(GGDC10S))

# This is the benchmark function (janitor 2.2.0 is used)
bmark <- function(...) {
  bench::mark(..., min_time = 0.01, check = FALSE) |> janitor::clean_names() |>
  fselect(expression, min, median, mem_alloc, n_itr, n_gc, total_time) |>
  fmutate(expression = names(expression)) |> dapply(as.character) |> qDF()
}

@

@@

<<faststatfun>>=
fmean(iris$Sepal.Length)
fmean(num_vars(iris))
identical(fmean(num_vars(iris)), fmean(as.matrix(num_vars(iris))))
fmean(iris$Sepal.Length, g = iris$Species)
fmean(num_vars(iris), g = iris$Species, nthreads = 4)
@

@@

<<TRA>>=
fmean(iris$Sepal.Length, g = iris$Species, TRA = "fill")[c(1:5, 51:55)]
@

@@

<<BY>>=
BY(num_vars(iris), g = iris$Species, FUN = mad) |> head(1)
@
<<fscale>>=
fscale(iris$Sepal.Length, g = iris$Species) |> fsd(iris$Species)
@

@@

<<wlddev_distinct>>=
fndistinct(wlddev)
@

@@

<<fsummarise>>=
wlddev |> fgroup_by(country, income) |>
  fsummarise(n = fnobs(LIFEEX),
             diff = fmax(LIFEEX) %-=% fmin(LIFEEX),
             mean_diff = fmean(fdiff(LIFEEX)),
             cor_PCGDP = pwcor(Dlog(LIFEEX), Dlog(PCGDP)),
             across(c(LIFEEX, PCGDP, POP), flast)) |>
  collap( ~ income, w = ~ POP, keep.w = FALSE) |> print(digits = 2)
@

@@

<<fsummarise_2>>=
wlddev |> fsubset(year >= 2015, income, PCGDP:POP) |>
  fgroup_by(income) |> fmean(POP)
@

@@

\pkg{collapse} thus offers several ways to reach the same outcome. This may be confusing, but all options can be distinguished in efficiency terms. It is thus useful to be mindful of computer resources and attempt to craft minimalistic solutions. For example, I recently combined multiple spatial datasets on points of interest (POIs) and needed to deduplicate them. I decided to keep the richest source for each location and POI type. After creating POI confidence, location, and type indicators comparable across sources (datasets), my deduplication expression ended up being a single line of the form \code{fsubset(data, source == fmode(source, list(location, type), confidence, "fill"))}---which retains POIs from the confidence-weighted most frequent (i.e., richest) source by location and type. This is very efficient because it avoids materializing intermediate datasets and relegates all computations to \fct{fmode}.
%
\subsection{Vectorizations for advanced tasks} \label{ssec:vfat}
%
\fct{fsummarise} and \fct{fmutate} follow an eager vectorization approach, implying that using any \emph{Fast Statistical Function} in an expression causes the whole expression to be vectorized. This only applies to visible expressions, \pkg{collapse} cannot read the contents of custom functions, thus such functions are always evaluated split-apply-combine. The eager vectorization approach enables efficient grouped calculation of more complex statistics. Below, I forecast the population of each region via linear regression (\code{POP ~ year}) in a fully vectorized way.
%
<<glinreg>>=
wlddev |> collap(POP ~ region + year, FUN = fsum) |>
  fmutate(POP = POP / 1e6) |> fgroup_by(region) |>
  fmutate(dmy = fmean(year, TRA = "-")) |>
  fsummarise(beta = fsum(POP, dmy) %/=% fsum(dmy, dmy),
             POP20 = flast(POP)) |>
  fmutate(POP21 = POP20 + beta, POP22 = POP21 + beta,
          POP23 = POP22 + beta, POP24 = POP23 + beta)
@
%
When \fct{fsummarise} evaluates an expression involving \emph{Fast Statistical Functions}, it sets their \code{g} argument with a grouping (\class{GRP}) object that is directly handed to C/C++, and also sets \code{use.g.names = FALSE}. Hence, weights (\code{w}) becomes the second positional argument. Similarly, \fct{fmutate} sets \code{g} and \code{TRA = "fill"}, which can be overwritten by the user (here with \code{TRA = "-"}). The expression \code{fsum(x, dmy) \%/=\% fsum(dmy, dmy)}, amounting to \code{cov(x, y)/var(y)}, is thus vectorized across groups and memory efficient---leveraging the weights (\code{w}) argument to \fct{fsum} to compute products (\code{v * dmy} and \code{dmy * dmy}) on the fly and division by reference (\code{\%/=\%}) to avoid an additional allocation. In a 2023 \href{https://sebkrantz.github.io/Rblog/2023/04/12/collapse-and-the-fastverse-reflecting-the-past-present-and-future/}{blog post}, I forecasted high-resolution population estimates for South Africa like this. Using 1 $km^2$ \href{https://www.worldpop.org/}{WorldPop} data available for years 2014-2020, running 1.6 million cell-regressions and obtaining 2 forecasts for 2021 and 2022 took only 0.3 seconds on my M1 Mac. Another neat example from the community, shared by Andrew Ghazi in a \href{https://andrewghazi.github.io/posts/collapse\_is\_sick/sick.html}{blog post}, vectorizes an expression to compute the $p$~value, \code{2 * pt(abs(fmean(x) * sqrt(6) / fsd(x)), 5, lower.tail = FALSE)}, across 300,000 groups for a simulation study, yielding a 70x performance increase over \pkg{dplyr}. \newline
% The eager vectorization approach of \pkg{collapse} here replaces \code{fmean(x)} and \code{fsd(x)} by their grouped versions and evaluates the entire expression once rather than 300k times as in \pkg{dplyr}. \newline

@@

<<wstats>>=
wlddev |> fsubset(is.finite(POP)) |> fgroup_by(region) |>
  fmutate(o = radixorder(GRPid(), LIFEEX)) |>
  fsummarise(min = fmin(LIFEEX),
             Q1 = fnth(LIFEEX, 0.25, POP, o = o, ties = "q8"),
             mean = fmean(LIFEEX, POP),
             median = fmedian(LIFEEX, POP, o = o),
             Q3 = fnth(LIFEEX, 0.75, POP, o = o, ties = "q8"),
             max = fmax(LIFEEX))
@

@@

<<GRP>>=
str(g <- GRP(wlddev, ~ income + OECD))
@

@@

<<gcomp>>=
add_vars(g$groups,
 get_vars(wlddev, "country") |> fmode(g, wlddev$POP, use = FALSE),
 get_vars(wlddev, c("PCGDP", "LIFEEX")) |> fmean(g, wlddev$POP, use = F),
 get_vars(wlddev, "POP") |> fsum(g, use = FALSE))
@

@@

% For advanced data aggregation, \pkg{collapse} also provides a convenience function, \fct{collap}, which, by default, uses \fct{fmean} for numeric, \fct{fmode} for categorical, and \fct{fsum} for weight columns, and preserves their order. The equivalent expression using this function would be
%
% <<collap>>=
% ollap(wlddev, country + PCGDP + LIFEEX ~ income + OECD, w = ~ POP)
% @
%
Similarly, data can be transformed, here using \fct{fwithin} to level average differences in economic status, adding back the overall mean after subtracting out group means:\footnote{\fct{add\_stub} adds a prefix (or suffix if \code{pre = FALSE}) to columns $\to$ \code{center\_PCGDP} and \code{center\_LIFEEX}.}
%
<<gtrans>>=
add_vars(wlddev) <- get_vars(wlddev, c("PCGDP", "LIFEEX")) |>
    fwithin(g, mean = "overall.mean") |> add_stub("center_")
@
%
% For (higher-dimensional) centering, \pkg{collapse} also has specialized function(s) \fct{f[hd]within} with additional options, and \fct{fscale} supports various scaling and centering operations.\newline

@@

<<fsumm_bench>>=
 bmark(collapse = collapse::fsummarise(mtcars, sum = sum(mpg)),
       dplyr = dplyr::summarise(mtcars, sum = sum(mpg)))
@

@@

<<exports>>=
set.seed(101)
exports <- expand.grid(y = 2001:2010, c = paste0("c", 1:10),
  s = paste0("s", 1:10)) |> fmutate(v = abs(rnorm(1e3))) |>
  colorder(c, s) |> fsubset(-sample.int(1e3, 500))
@

@@

<<ts_example_2>>=
.c(y, v) %=% fsubset(exports, c == "c1" & s == "s7", -c, -s)
print(y)
fgrowth(v, t = y) |> round(2)
fgrowth(v, -1:3, t = y) |> head(4)
@

@@

<<example_growth>>=
G(exports, -1:2, by = v ~ c + s, t = ~ y) |> head(3)
tfm(exports, fgrowth(list(v = v), -1:2, g = list(c, s), t = y)) |> head(3)
@

@@

<<example_growth_continued>>=
A <- exports |> fgroup_by(c, s) |> fmutate(gv = G(v, t = y)) |> fungroup()
head(B <- exports |> fmutate(gv = G(v, g = list(c, s), t = y)), 4)
identical(A, B)
@

@@

An example using the \code{exports} data follows. Note that data need not be sorted for indexing.
<<indexing>>=
exportsi <- exports |> findex_by(c, s, y)
exportsi |> G() |> head(5)
exportsi |> findex() |> print(2)
@
The index statistics are: \code{[N. ids] | [N. periods (total periods: (max-min)/GCD)]}. % This extracts an \class{indexes\_series} and demonstrates its properties.
<<indexing_2>>=
vi <- exportsi$v; str(vi, width = 70, strict = "cut")
is_irregular(vi)
vi |> psmat() |> head(3)
fdiff(vi) |> psmat() |> head(3)
@
\fct{psmat}, for panel-series to matrix, generates a matrix/array from panel data. Thanks to deep indexation, indexed computations work in arbitrary data masking environments:
<<indexing_3>>=
settransform(exportsi, v_ld = Dlog(v))
lm(v_ld ~ L(v_ld, 1:2), exportsi) |> summary() |> coef() |> round(3)
@
Indexed series/frames also support transformation such as grouped scaling with \fct{fscale} or demeaning with \fct{fwithin}. Functions \fct{psacf}/\fct{pspacf}/\fct{psccf} provide panel-data autocorrelation functions, which are computed using group-scaled and suitably lagged panel-series. The \class{index\_df} attached to these objects can also be used with other general tools such as \code{collapse::BY()} to perform grouped computations using 3rd-party functions. An example of calculating a 5-year rolling average is given below (\fct{ix} abbreviates \fct{findex}).
%
<<rollmean>>=
BY(vi, ix(vi)$c.s, data.table::frollmean, 5) |> head(10)
@
% Last but not least, the performance of these classes is second to none.\footnote{See the \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{useR 2022 presentation} slide 40 for a small benchmark.}
%
\section{Table joins and pivots} \label{sec:join_pivot}
%
Among all \href{https://sebkrantz.github.io/collapse/reference/fast-data-manipulation.html}{data manipulation functions} \pkg{collapse} provides, its implementations of table \href{https://sebkrantz.github.io/collapse/reference/join.html}{joins} and \href{https://sebkrantz.github.io/collapse/reference/pivot.html}{pivots} are particularly noteworthy since they offer several new features, including rich verbosity for table joins, pivots supporting variable labels, and recast pivots. Both implementations provide outstanding computational performance, syntax, and memory efficiency.
%
\subsection{Joins}
%
Compared to commercial software such as \proglang{STATA} \citep{STATA}, the implementation of joins in most open-source software provides no information on how many records were joined from both tables. This often provokes manual efforts to validate the join operation. \fct{collapse::join} provides many options to understand table join operations. Its syntax is:
\begin{Code}
join(x, y, on = NULL, how = "left", suffix = NULL, validate = "m:m",
  multiple = FALSE, sort = FALSE, keep.col.order = TRUE,
  drop.dup.cols = FALSE, verbose = 1, column = NULL, attr = NULL, ...)
\end{Code}
It defaults to left join and only takes first matches from \code{y} (\code{multiple = FALSE}), thus, it simply adds columns to \code{x}, which is efficient and sufficient/desired in many cases. By default (\code{verbose = 1}), it prints information about the join operation and number of records joined. \newline

>>>>

%% For Sweave-based articles about R packages:
%% need no \usepackage{Sweave}
%% need no \usepackage{Sweave.sty}

% \renewenvironment{knitrout}{\setlength{\topsep}{0mm}}{}

@@

\begin{Schunk}
\begin{Sinput}
R> fmean(iris$Sepal.Length)
\end{Sinput}
\begin{Soutput}
[1] 5.843
\end{Soutput}
\begin{Sinput}
R> fmean(num_vars(iris))
\end{Sinput}
\begin{Soutput}
Sepal.Length  Sepal.Width Petal.Length  Petal.Width 
       5.843        3.057        3.758        1.199 
\end{Soutput}
\begin{Sinput}
R> identical(fmean(num_vars(iris)), fmean(as.matrix(num_vars(iris))))
\end{Sinput}
\begin{Soutput}
[1] TRUE
\end{Soutput}
\begin{Sinput}
R> fmean(iris$Sepal.Length, g = iris$Species)
\end{Sinput}
\begin{Soutput}
    setosa versicolor  virginica 
     5.006      5.936      6.588 
\end{Soutput}
\begin{Sinput}
R> fmean(num_vars(iris), g = iris$Species, nthreads = 4)
\end{Sinput}
\begin{Soutput}
           Sepal.Length Sepal.Width Petal.Length Petal.Width
setosa            5.006       3.428        1.462       0.246
versicolor        5.936       2.770        4.260       1.326
virginica         6.588       2.974        5.552       2.026
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> fmean(iris$Sepal.Length, g = iris$Species, TRA = "fill")[c(1:5, 51:55)]
\end{Sinput}
\begin{Soutput}
 [1] 5.006 5.006 5.006 5.006 5.006 5.936 5.936 5.936 5.936 5.936
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> BY(num_vars(iris), g = iris$Species, FUN = mad) |> head(1)
\end{Sinput}
\begin{Soutput}
       Sepal.Length Sepal.Width Petal.Length Petal.Width
setosa       0.2965      0.3706       0.1483           0
\end{Soutput}
\end{Schunk}
\begin{Schunk}
\begin{Sinput}
R> fscale(iris$Sepal.Length, g = iris$Species) |> fsd(iris$Species)
\end{Sinput}
\begin{Soutput}
    setosa versicolor  virginica 
         1          1          1 
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> fndistinct(wlddev)
\end{Sinput}
\begin{Soutput}
country   iso3c    date    year  decade  region  income    OECD   PCGDP 
    216     216      61      61       7       7       4       2    9470 
 LIFEEX    GINI     ODA     POP 
  10548     368    7832   12877 
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> wlddev |> fgroup_by(country, income) |>
+    fsummarise(n = fnobs(LIFEEX),
+               diff = fmax(LIFEEX) %-=% fmin(LIFEEX),
+               mean_diff = fmean(fdiff(LIFEEX)),
+               cor_PCGDP = pwcor(Dlog(LIFEEX), Dlog(PCGDP)),
+               across(c(LIFEEX, PCGDP, POP), flast)) |>
+    collap( ~ income, w = ~ POP, keep.w = FALSE) |> print(digits = 2)
\end{Sinput}
\begin{Soutput}
        country              income  n diff mean_diff cor_PCGDP LIFEEX PCGDP
1 United States         High income 60   13      0.23  -0.00828     81 44599
2      Ethiopia          Low income 60   26      0.43   0.27153     64   691
3         India Lower middle income 60   24      0.41  -0.15366     69  2324
4         China Upper middle income 60   27      0.45   0.00077     76  8749
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> wlddev |> fsubset(year >= 2015, income, PCGDP:POP) |>
+    fgroup_by(income) |> fmean(POP)
\end{Sinput}
\begin{Soutput}
               income   sum.POP PCGDP LIFEEX  GINI        ODA
1         High income 5.901e+09 43340  80.70 36.14   81398948
2          Low income 3.296e+09   663  63.05 39.13 2466732035
3 Lower middle income 1.490e+10  2177  68.31 36.48 2234540565
4 Upper middle income 1.318e+10  8168  75.51 41.68  -86630117
\end{Soutput}
\end{Schunk}

@@

\pkg{collapse} thus offers several ways to reach the same outcome. This may be confusing, but all options can be distinguished in efficiency terms. It is thus useful to be mindful of computer resources and attempt to craft minimalistic solutions. For example, I recently combined multiple spatial datasets on points of interest (POIs) and needed to deduplicate them. I decided to keep the richest source for each location and POI type. After creating POI confidence, location, and type indicators comparable across sources (datasets), my deduplication expression ended up being a single line of the form \code{fsubset(data, source == fmode(source, list(location, type), confidence, "fill"))}---which retains POIs from the confidence-weighted most frequent (i.e., richest) source by location and type. This is very efficient because it avoids materializing intermediate datasets and relegates all computations to \fct{fmode}.
\subsection{Vectorizations for advanced tasks} \label{ssec:vfat}
\fct{fsummarise} and \fct{fmutate} follow an eager vectorization approach, implying that using any \emph{Fast Statistical Function} in an expression causes the whole expression to be vectorized. This only applies to visible expressions, \pkg{collapse} cannot read the contents of custom functions, thus such functions are always evaluated split-apply-combine. The eager vectorization approach enables efficient grouped calculation of more complex statistics. Below, I forecast the population of each region via linear regression (\code{POP ~ year}) in a fully vectorized way.
\begin{Schunk}
\begin{Sinput}
R> wlddev |> collap(POP ~ region + year, FUN = fsum) |>
+    fmutate(POP = POP / 1e6) |> fgroup_by(region) |>
+    fmutate(dmy = fmean(year, TRA = "-")) |>
+    fsummarise(beta = fsum(POP, dmy) %/=% fsum(dmy, dmy),
+               POP20 = flast(POP)) |>
+    fmutate(POP21 = POP20 + beta, POP22 = POP21 + beta,
+            POP23 = POP22 + beta, POP24 = POP23 + beta)
\end{Sinput}
\begin{Soutput}
                      region   beta  POP20  POP21  POP22  POP23  POP24
1        East Asia & Pacific 18.731 2291.4 2310.2 2328.9 2347.6 2366.4
2      Europe & Central Asia  2.463  921.2  923.7  926.1  928.6  931.0
3  Latin America & Caribbean  6.460  646.4  652.9  659.4  665.8  672.3
4 Middle East & North Africa  5.409  456.7  462.1  467.5  472.9  478.3
5              North America  2.301  365.9  368.2  370.5  372.8  375.1
6                 South Asia 19.640 1835.8 1855.4 1875.1 1894.7 1914.3
7         Sub-Saharan Africa 12.852 1103.5 1116.3 1129.2 1142.0 1154.9
\end{Soutput}
\end{Schunk}
When \fct{fsummarise} evaluates an expression involving \emph{Fast Statistical Functions}, it sets their \code{g} argument with a grouping (\class{GRP}) object that is directly handed to C/C++, and also sets \code{use.g.names = FALSE}. Hence, weights (\code{w}) becomes the second positional argument. Similarly, \fct{fmutate} sets \code{g} and \code{TRA = "fill"}, which can be overwritten by the user (here with \code{TRA = "-"}). The expression \code{fsum(x, dmy) \%/=\% fsum(dmy, dmy)}, amounting to \code{cov(x, y)/var(y)}, is thus vectorized across groups and memory efficient---leveraging the weights (\code{w}) argument to \fct{fsum} to compute products (\code{v * dmy} and \code{dmy * dmy}) on the fly and division by reference (\code{\%/=\%}) to avoid an additional allocation. In a 2023 \href{https://sebkrantz.github.io/Rblog/2023/04/12/collapse-and-the-fastverse-reflecting-the-past-present-and-future/}{blog post}, I forecasted high-resolution population estimates for South Africa like this. Using 1 $km^2$ \href{https://www.worldpop.org/}{WorldPop} data available for years 2014-2020, running 1.6 million cell-regressions and obtaining 2 forecasts for 2021 and 2022 took only 0.3 seconds on my M1 Mac. Another neat example from the community, shared by Andrew Ghazi in a \href{https://andrewghazi.github.io/posts/collapse\_is\_sick/sick.html}{blog post}, vectorizes an expression to compute the $p$~value, \code{2 * pt(abs(fmean(x) * sqrt(6) / fsd(x)), 5, lower.tail = FALSE)}, across 300,000 groups for a simulation study, yielding a 70x performance increase over \pkg{dplyr}. \newline
% The eager vectorization approach of \pkg{collapse} here replaces \code{fmean(x)} and \code{fsd(x)} by their grouped versions and evaluates the entire expression once rather than 300k times as in \pkg{dplyr}. \newline

@@

\begin{Schunk}
\begin{Sinput}
R> wlddev |> fsubset(is.finite(POP)) |> fgroup_by(region) |>
+    fmutate(o = radixorder(GRPid(), LIFEEX)) |>
+    fsummarise(min = fmin(LIFEEX),
+               Q1 = fnth(LIFEEX, 0.25, POP, o = o, ties = "q8"),
+               mean = fmean(LIFEEX, POP),
+               median = fmedian(LIFEEX, POP, o = o),
+               Q3 = fnth(LIFEEX, 0.75, POP, o = o, ties = "q8"),
+               max = fmax(LIFEEX))
\end{Sinput}
\begin{Soutput}
                      region   min    Q1  mean median    Q3   max
1        East Asia & Pacific 18.91 65.28 68.45  69.67 73.86 85.08
2      Europe & Central Asia 45.37 68.68 72.30  71.58 76.67 85.42
3  Latin America & Caribbean 41.76 65.17 69.16  70.87 74.48 82.19
4 Middle East & North Africa 29.92 61.96 66.65  69.12 72.64 82.80
5              North America 68.90 73.57 75.54  75.62 78.38 82.05
6                 South Asia 32.45 55.08 60.19  62.00 66.67 78.92
7         Sub-Saharan Africa 26.17 46.51 52.53  52.23 58.32 74.51
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> str(g <- GRP(wlddev, ~ income + OECD))
\end{Sinput}
\begin{Soutput}
Class 'GRP'  hidden list of 9
 $ N.groups    : int 6
 $ group.id    : int [1:13176] 3 3 3 3 3 3 3 3 3 3 ...
 $ group.sizes : int [1:6] 2745 2074 1830 2867 3538 122
 $ groups      :'data.frame':	6 obs. of  2 variables:
  ..$ income: Factor w/ 4 levels "High income",..: 1 1 2 3 4 4
  .. ..- attr(*, "label")= chr "Income Level"
  ..$ OECD  : logi [1:6] FALSE TRUE FALSE FALSE FALSE TRUE
  .. ..- attr(*, "label")= chr "Is OECD Member Country?"
 $ group.vars  : chr [1:2] "income" "OECD"
 $ ordered     : Named logi [1:2] TRUE FALSE
  ..- attr(*, "names")= chr [1:2] "ordered" "sorted"
 $ order       : int [1:13176] 245 246 247 248 249 250 251 252 253 254 ...
  ..- attr(*, "starts")= int [1:6] 1 2746 4820 6650 9517 13055
  ..- attr(*, "maxgrpn")= int 3538
  ..- attr(*, "sorted")= logi FALSE
 $ group.starts: int [1:6] 245 611 1 306 62 7687
 $ call        : language GRP.default(X = wlddev, by = ~income + OECD)
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> add_vars(g$groups,
+   get_vars(wlddev, "country") |> fmode(g, wlddev$POP, use = FALSE),
+   get_vars(wlddev, c("PCGDP", "LIFEEX")) |> fmean(g, wlddev$POP, use = F),
+   get_vars(wlddev, "POP") |> fsum(g, use = FALSE))
\end{Sinput}
\begin{Soutput}
               income  OECD       country   PCGDP LIFEEX       POP
1         High income FALSE  Saudi Arabia 22426.7  73.00 3.114e+09
2         High income  TRUE United States 31749.6  75.84 5.573e+10
3          Low income FALSE      Ethiopia   557.1  53.51 2.095e+10
4 Lower middle income FALSE         India  1238.8  60.59 1.138e+11
5 Upper middle income FALSE         China  3820.6  68.21 1.114e+11
6 Upper middle income  TRUE        Mexico  8311.2  69.06 8.162e+09
\end{Soutput}
\end{Schunk}

@@

% For advanced data aggregation, \pkg{collapse} also provides a convenience function, \fct{collap}, which, by default, uses \fct{fmean} for numeric, \fct{fmode} for categorical, and \fct{fsum} for weight columns, and preserves their order. The equivalent expression using this function would be
%
% <<collap>>=
% ollap(wlddev, country + PCGDP + LIFEEX ~ income + OECD, w = ~ POP)
% @
%
Similarly, data can be transformed, here using \fct{fwithin} to level average differences in economic status, adding back the overall mean after subtracting out group means:\footnote{\fct{add\_stub} adds a prefix (or suffix if \code{pre = FALSE}) to columns $\to$ \code{center\_PCGDP} and \code{center\_LIFEEX}.}
%
\begin{Schunk}
\begin{Sinput}
R> add_vars(wlddev) <- get_vars(wlddev, c("PCGDP", "LIFEEX")) |>
+      fwithin(g, mean = "overall.mean") |> add_stub("center_")
\end{Sinput}
\end{Schunk}
%
% For (higher-dimensional) centering, \pkg{collapse} also has specialized function(s) \fct{f[hd]within} with additional options, and \fct{fscale} supports various scaling and centering operations.\newline

@@

\begin{Schunk}
\begin{Sinput}
R>  bmark(collapse = collapse::fsummarise(mtcars, sum = sum(mpg)),
+         dplyr = dplyr::summarise(mtcars, sum = sum(mpg)))
\end{Sinput}
\begin{Soutput}
  expression      min   median mem_alloc n_itr n_gc total_time
1   collapse   3.61µs   4.26µs   14.33KB  9999    1    43.81ms
2      dplyr 211.27µs 240.26µs    3.92MB  7651   52      1.86s
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> set.seed(101)
R> exports <- expand.grid(y = 2001:2010, c = paste0("c", 1:10),
+    s = paste0("s", 1:10)) |> fmutate(v = abs(rnorm(1e3))) |>
+    colorder(c, s) |> fsubset(-sample.int(1e3, 500))
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> .c(y, v) %=% fsubset(exports, c == "c1" & s == "s7", -c, -s)
R> print(y)
\end{Sinput}
\begin{Soutput}
[1] 2001 2002 2004 2005 2007 2008 2009 2010
\end{Soutput}
\begin{Sinput}
R> fgrowth(v, t = y) |> round(2)
\end{Sinput}
\begin{Soutput}
[1]     NA 175.52     NA -22.37     NA 624.27 -79.01 534.56
\end{Soutput}
\begin{Sinput}
R> fgrowth(v, -1:3, t = y) |> head(4)
\end{Sinput}
\begin{Soutput}
        FG1     --     G1   L2G1   L3G1
[1,] -63.71 0.3893     NA     NA     NA
[2,]     NA 1.0726 175.52     NA     NA
[3,]  28.82 0.8450     NA -21.22 117.05
[4,]     NA 0.6559 -22.37     NA -38.85
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> G(exports, -1:2, by = v ~ c + s, t = ~ y) |> head(3)
\end{Sinput}
\begin{Soutput}
   c  s    y  FG1.v      v   G1.v L2G1.v
1 c1 s1 2002 -18.15 0.5525     NA     NA
2 c1 s1 2003 214.87 0.6749  22.17     NA
3 c1 s1 2004 -31.02 0.2144 -68.24  -61.2
\end{Soutput}
\begin{Sinput}
R> tfm(exports, fgrowth(list(v = v), -1:2, g = list(c, s), t = y)) |> head(3)
\end{Sinput}
\begin{Soutput}
   c  s    y      v  FG1.v   G1.v L2G1.v
1 c1 s1 2002 0.5525 -18.15     NA     NA
2 c1 s1 2003 0.6749 214.87  22.17     NA
3 c1 s1 2004 0.2144 -31.02 -68.24  -61.2
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> A <- exports |> fgroup_by(c, s) |> fmutate(gv = G(v, t = y)) |> fungroup()
R> head(B <- exports |> fmutate(gv = G(v, g = list(c, s), t = y)), 4)
\end{Sinput}
\begin{Soutput}
   c  s    y      v     gv
1 c1 s1 2002 0.5525     NA
2 c1 s1 2003 0.6749  22.17
3 c1 s1 2004 0.2144 -68.24
4 c1 s1 2005 0.3108  44.98
\end{Soutput}
\begin{Sinput}
R> identical(A, B)
\end{Sinput}
\begin{Soutput}
[1] TRUE
\end{Soutput}
\end{Schunk}

@@

An example using the \code{exports} data follows. Note that data need not be sorted for indexing.
\begin{Schunk}
\begin{Sinput}
R> exportsi <- exports |> findex_by(c, s, y)
R> exportsi |> G() |> head(5)
\end{Sinput}
\begin{Soutput}
   c  s    y   G1.v
1 c1 s1 2002     NA
2 c1 s1 2003  22.17
3 c1 s1 2004 -68.24
4 c1 s1 2005  44.98
5 c1 s1 2006 277.76

Indexed by:  c.s [1] | y [5 (10)] 
\end{Soutput}
\begin{Sinput}
R> exportsi |> findex() |> print(2)
\end{Sinput}
\begin{Soutput}
    c.s    y
1 c1.s1 2002
2 c1.s1 2003
---                
499 c10.s10 2007
500 c10.s10 2009

c.s [100] | y [10]
\end{Soutput}
\end{Schunk}
The index statistics are: \code{[N. ids] | [N. periods (total periods: (max-min)/GCD)]}. % This extracts an \class{indexes\_series} and demonstrates its properties.
\begin{Schunk}
\begin{Sinput}
R> vi <- exportsi$v; str(vi, width = 70, strict = "cut")
\end{Sinput}
\begin{Soutput}
 'indexed_series' num [1:500] 0.552 0.675 0.214 0.311 1.174 ...
 - attr(*, "index_df")=Classes 'index_df', 'pindex' and 'data.frame'..
  ..$ c.s: Factor w/ 100 levels "c1.s1","c2.s1",..: 1 1 1 1 1 1 1 1 ..
  ..$ y  : Ord.factor w/ 10 levels "2001"<"2002"<..: 2 3 4 5 6 7 8 9..
  ..- attr(*, "nam")= chr [1:3] "c" "s" "y"
\end{Soutput}
\begin{Sinput}
R> is_irregular(vi)
\end{Sinput}
\begin{Soutput}
[1] TRUE
\end{Soutput}
\begin{Sinput}
R> vi |> psmat() |> head(3)
\end{Sinput}
\begin{Soutput}
      2001  2002  2003  2004  2005 2006  2007   2008  2009  2010
c1.s1   NA 0.552 0.675 0.214 0.311 1.17 0.619 0.1127 0.917 0.223
c2.s1   NA 0.795    NA    NA 0.237   NA    NA 0.0585 0.818    NA
c3.s1   NA 0.709 0.268 1.464    NA   NA 0.467 0.1193 0.467    NA
\end{Soutput}
\begin{Sinput}
R> fdiff(vi) |> psmat() |> head(3)
\end{Sinput}
\begin{Soutput}
      2001 2002   2003   2004   2005  2006   2007   2008  2009   2010
c1.s1   NA   NA  0.122 -0.461 0.0964 0.863 -0.555 -0.506 0.804 -0.694
c2.s1   NA   NA     NA     NA     NA    NA     NA     NA 0.759     NA
c3.s1   NA   NA -0.441  1.196     NA    NA     NA -0.348 0.348     NA
\end{Soutput}
\end{Schunk}
\fct{psmat}, for panel-series to matrix, generates a matrix/array from panel data. Thanks to deep indexation, indexed computations work in arbitrary data masking environments:
\begin{Schunk}
\begin{Sinput}
R> settransform(exportsi, v_ld = Dlog(v))
R> lm(v_ld ~ L(v_ld, 1:2), exportsi) |> summary() |> coef() |> round(3)
\end{Sinput}
\begin{Soutput}
               Estimate Std. Error t value Pr(>|t|)
(Intercept)      -0.008      0.141  -0.058    0.954
L(v_ld, 1:2)L1   -0.349      0.115  -3.042    0.004
L(v_ld, 1:2)L2   -0.033      0.154  -0.215    0.831
\end{Soutput}
\end{Schunk}
Indexed series/frames also support transformation such as grouped scaling with \fct{fscale} or demeaning with \fct{fwithin}. Functions \fct{psacf}/\fct{pspacf}/\fct{psccf} provide panel-data autocorrelation functions, which are computed using group-scaled and suitably lagged panel-series. The \class{index\_df} attached to these objects can also be used with other general tools such as \code{collapse::BY()} to perform grouped computations using 3rd-party functions. An example of calculating a 5-year rolling average is given below (\fct{ix} abbreviates \fct{findex}).
%
\begin{Schunk}
\begin{Sinput}
R> BY(vi, ix(vi)$c.s, data.table::frollmean, 5) |> head(10)
\end{Sinput}
\begin{Soutput}
 [1]     NA     NA     NA     NA 0.5853 0.5986 0.4861 0.6267 0.6092     NA
